---
title: "General Conference NLP Report"
subtitle: "MiniLM (+ optional subword TF–IDF pooling) & LDA across the full corpus"
author: "conferencestats analysis"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
    number_sections: true
    df_print: paged
    fig_width: 10
    fig_height: 5.5
params:
  repo_root: !r if (nzchar(Sys.getenv("CONFERENCESTATS_ROOT", ""))) Sys.getenv("CONFERENCESTATS_ROOT") else normalizePath("..")
  lda_k: 16
  minilm_batch_size: 48
  # When TRUE: last-layer hidden states are pooled with weights ∝ subword tf×idf (same tokenizer as MiniLM).
  use_subword_tfidf_pool: true
  minilm_max_length: 256
  tfidf_pool_batch_size: 8
  max_paragraph_rows: null
  min_paragraph_chars: 40
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.path = "figure_gc_nlp/",
  dpi = 120
)
repo_root <- params$repo_root
if (!dir.exists(file.path(repo_root, "R"))) {
  stop("params$repo_root must point at the conferencestats package root.")
}
```

# Executive summary

This report uses **`sentence-transformers/all-MiniLM-L6-v2`** (384-d, same WordPiece tokenizer as the stock encoder) via **`reticulate`**, on **every paragraph** aligned to **`generalconference`**, to track **prescriptive vs invitational** semantic poles from regex-mined prototypes. With **`params$use_subword_tfidf_pool: true`** (default), paragraph vectors come from **last-layer subword positions weighted by normalized tf×idf** on token IDs (then L2-normalized)—not the library’s default mean pooling. It then fits **LDA** on **every talk** and plots **topic prevalence over year**.

> **Note:** Use **raw paragraph strings**. Optional `params$use_subword_tfidf_pool: false` restores plain **`SentenceTransformer.encode()`** mean pooling (requires `sentence_transformers`).

```{r branding, echo=FALSE, results='asis'}
cat(
  '<blockquote style="border-left:4px solid #2c5282;padding-left:1em;color:#2d3748;">',
  "<strong>Corpus:</strong> <code>generalconference::genconf</code> (flattened). ",
  "<br/><strong>MiniLM:</strong> ",
  '<a href="https://huggingface.co/sentence-transformers/all-MiniLM-L6-v2">all-MiniLM-L6-v2</a>.',
  "</blockquote>"
)
```

# Prerequisites

Install once (R):

```r
install.packages(c(
  "pkgload", "rmarkdown", "knitr", "reticulate", "dplyr", "tidyr",
  "stringr", "purrr", "ggplot2", "scales", "mgcv", "quanteda",
  "topicmodels", "tibble", "remotes", "rprojroot"
))
remotes::install_github("bryanwhiting/generalconference@v0.3.2")
```

Python (CPU is fine; first run downloads the model from HuggingFace):

```{r py-bootstrap, eval=FALSE}
reticulate::miniconda_install() # optional if you want an isolated Python
reticulate::py_install(
  c("transformers", "sentence-transformers", "torch", "numpy"),
  pip = TRUE
)
```

```{r py-ensure}
library(reticulate)
use_pool <- isTRUE(params$use_subword_tfidf_pool)
have_st <- tryCatch(
  {
    import("sentence_transformers")
    TRUE
  },
  error = function(e) FALSE
)
have_tf <- tryCatch(
  {
    import("transformers")
    TRUE
  },
  error = function(e) FALSE
)
if (use_pool) {
  if (!have_tf) {
    message("Installing transformers + torch via pip (one-time, can take several minutes)...")
    py_install(c("transformers", "torch"), pip = TRUE)
  }
} else if (!have_st) {
  message("Installing sentence-transformers + torch via pip (one-time, can take several minutes)...")
  py_install(c("sentence-transformers", "torch"), pip = TRUE)
}
```

# Load package and full talk table

```{r load-pkg}
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(scales)
library(mgcv)
library(quanteda)
library(topicmodels)
library(tibble)
pkgload::load_all(repo_root, quiet = TRUE)

if (!requireNamespace("generalconference", quietly = TRUE)) {
  stop("Install generalconference v0.3.2 (see Prerequisites).")
}

talks <- gc_normalize_talks(gc_as_tibble_generalconference())
stopifnot(nrow(talks) > 100L)

talks_summary <- talks |>
  summarise(
    n_talks = dplyr::n(),
    year_min = min(year, na.rm = TRUE),
    year_max = max(year, na.rm = TRUE),
    n_words = sum(word_count, na.rm = TRUE)
  )
knitr::kable(talks_summary, caption = "Full corpus summary")
```

# Part A — Paragraphs & MiniLM embeddings

## Build paragraph table

```{r paragraphs}
para <- gc_as_paragraph_tibble_generalconference() |>
  inner_join(
    talks |>
      select(talk_id, year, month, conference_date, speaker, title, url, word_count),
    by = c("year", "month", "speaker", "title", "url")
  ) |>
  filter(nchar(.data$text) >= params$min_paragraph_chars)

if (!is.null(params$max_paragraph_rows) && is.numeric(params$max_paragraph_rows) &&
  params$max_paragraph_rows > 0L) {
  set.seed(42)
  para <- dplyr::slice_sample(para, n = min(nrow(para), as.integer(params$max_paragraph_rows)))
}

knitr::kable(
  utils::head(as.data.frame(para), 4),
  caption = "Sample paragraph rows"
)
message("Paragraphs for MiniLM: ", nrow(para))
```

## Regex-mined prototype spans (same spirit as prior analysis)

```{r regex-bundles}
presc_rx <- regex(
  paste(
    "do\\s+not",
    "don'?t",
    "can\\s+not",
    "cannot",
    "must\\s+not",
    "should\\s+not",
    "ought\\s+not",
    "thou\\s+shalt\\s+not",
    "shall\\s+not",
    "never\\s+\\w+",
    "cease\\s+(from|to)",
    "beware",
    "wo\\s*(e)?\\s*unto",
    "woe\\s+unto",
    "condemn",
    "rebellion\\s+against",
    "must\\s+stop",
    sep = "|"
  ),
  ignore_case = TRUE
)

gentle_rx <- regex(
  paste(
    "i invite",
    "we invite",
    "invitation",
    "come\\s+unto\\s+christ",
    "jesus\\s+christ",
    "our\\s+savior",
    "redeemer",
    "his\\s+love",
    "peace\\s+in\\s+christ",
    "gentle",
    "tender",
    "mercies",
    "consider\\s+how",
    "ponder",
    sep = "|"
  ),
  ignore_case = TRUE
)

para <- para |>
  mutate(
    has_presc = str_detect(.data$text, presc_rx),
    has_gentle = str_detect(.data$text, gentle_rx)
  )
```

## Encode all paragraphs with MiniLM

```{r minilm-encode}
py_dir <- normalizePath(file.path(repo_root, "analysis", "python"))
sys <- reticulate::import("sys", convert = FALSE)
sys$path$insert(0L, py_dir)
txt <- as.character(para$text)

if (isTRUE(params$use_subword_tfidf_pool)) {
  stw <- reticulate::import("st_minilm_tfidf_pool", convert = TRUE)
  ml <- as.integer(params$minilm_max_length)
  tbs <- params$tfidf_pool_batch_size
  bs_pool <- if (is.numeric(tbs) && length(tbs) == 1L && !is.na(tbs) && tbs > 0) {
    as.integer(tbs)
  } else {
    as.integer(min(16L, params$minilm_batch_size))
  }
  idf <- stw$compute_subword_idf(txt, max_length = ml)
  emb <- stw$encode_tfidf_weighted_pool(txt, idf, max_length = ml, batch_size = bs_pool)
  mod <- stw
} else {
  st <- reticulate::import("st_minilm", convert = TRUE)
  emb <- st$encode_texts(txt, batch_size = as.integer(params$minilm_batch_size))
  mod <- st
}

stopifnot(is.matrix(emb), nrow(emb) == nrow(para), ncol(emb) == 384L)

u_presc <- mod$centroid_unit(emb, para$has_presc)
u_gent <- mod$centroid_unit(emb, para$has_gentle)
if (is.null(u_presc) || is.null(u_gent)) {
  stop("Could not build MiniLM prototypes — broaden regex or corpus.")
}

cos_p <- as.numeric(mod$cosine_to_rows(emb, u_presc))
cos_g <- as.numeric(mod$cosine_to_rows(emb, u_gent))

para_scored <- para |>
  mutate(
    cos_presc = cos_p,
    cos_gentle = cos_g,
    net_presc = .data$cos_presc - .data$cos_gentle
  )
```

## Talk- and year-level aggregation + plots

```{r minilm-aggregates}
talk_sc <- para_scored |>
  group_by(talk_id, year, month, conference_date, speaker) |>
  summarise(
    n_para = dplyr::n(),
    mean_net = mean(.data$net_presc, na.rm = TRUE),
    q75_net = quantile(.data$net_presc, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

m_net <- mgcv::gam(mean_net ~ s(year), data = talk_sc, method = "REML")
```

```{r plot-minilm-year, fig.cap="**MiniLM:** distribution of talk-level mean (cos prescriptive prot. − cos gentle prot.) by year. Trend = GAM smooth."}
ggplot(talk_sc, aes(year, mean_net)) +
  geom_hline(yintercept = 0, linetype = 3, alpha = 0.5) +
  geom_point(aes(size = n_para), alpha = 0.12, color = "#1a365d") +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 12), color = "#c53030", fill = "#c5303033") +
  scale_size(range = c(0.4, 3), guide = "none") +
  labs(
    title = expression(bold("MiniLM") ~ "prescriptive vs. gentle prototype (full paragraph corpus)"),
    subtitle = "Per talk: mean paragraph-level net semantic alignment | sentence-transformers/all-MiniLM-L6-v2",
    x = "Conference year",
    y = "Mean net cosine (prescriptive − gentle)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

```{r plot-minilm-decade, fig.cap="Decadal means with standard error."}
by_dec <- talk_sc |>
  mutate(decade = floor(year / 10) * 10) |>
  group_by(decade) |>
  summarise(
    n = dplyr::n(),
    m = mean(mean_net),
    se = sd(mean_net) / sqrt(dplyr::n()),
    .groups = "drop"
  )

ggplot(by_dec, aes(decade, m)) +
  geom_ribbon(aes(ymin = m - se, ymax = m + se), alpha = 0.25, fill = "#2b6cb0") +
  geom_line(color = "#2c5282", linewidth = 0.9) +
  geom_point(size = 2.5, color = "#1a365d") +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  labs(
    title = "Decadal compression of the MiniLM prescriptive index",
    x = "Decade",
    y = "Mean net cosine ± SE"
  ) +
  theme_minimal(base_size = 13)
```

```{r minilm-gam-note, echo=FALSE, results='asis'}
cat(
  "<p><strong>GAM</strong> <code>mean_net ~ s(year)</code>: edf = ",
  format(sum(m_net$edf[-1]), digits = 2),
  ", approximate p-value = ",
  format.pval(summary(m_net)$s.table[1, 4], digits = 3),
  ".</p>"
)
```

# Part B — LDA on **every talk**

```{r lda-corpus}
corp_lda <- corpus(
  talks,
  text_field = "text",
  docid_field = "talk_id"
)

toks_lda <- tokens(corp_lda, remove_punct = TRUE, remove_numbers = FALSE) |>
  tokens_remove(stopwords("en")) |>
  tokens_wordstem(language = "en")

dfm_lda <- dfm(toks_lda) |>
  dfm_trim(
    min_docfreq = 5,
    max_docfreq = 0.5,
    docfreq_type = "prop"
  )

message("LDA DFM: ", quanteda::ndoc(dfm_lda), " docs × ", quanteda::nfeat(dfm_lda), " feats")
```

```{r lda-fit}
set.seed(42)
dtm <- convert(dfm_lda, to = "topicmodels")
lda <- LDA(dtm, k = as.integer(params$lda_k), control = list(seed = 42, verbose = 0))
gamma <- posterior(lda)$topics
 colnames(gamma) <- paste0("topic_", seq_len(ncol(gamma)))

tt <- topicmodels::terms(lda, 12)
## tt: rows = ranked terms, cols = topics — transpose so each row is one topic
top_terms <- as.data.frame(t(tt), stringsAsFactors = FALSE)
colnames(top_terms) <- paste0("w", seq_len(ncol(top_terms)))
top_terms <- dplyr::bind_cols(
  topic = paste0("topic_", seq_len(nrow(top_terms))),
  top_terms
)
knitr::kable(
  utils::head(top_terms, 8),
  caption = "Top stemmed tokens per topic (one row per topic; columns are ranked terms)"
)
```

```{r lda-year-panel}
doc_ids <- docnames(dfm_lda)
## dfm_trim drops empty documents; row order of gamma matches doc_ids
if (!all(doc_ids %in% talks$talk_id)) {
  stop("LDA docnames missing from talks — check corpus alignment.")
}

gamma_df <- as_tibble(gamma) |>
  mutate(talk_id = doc_ids) |>
  left_join(talks |> select(talk_id, year), by = "talk_id") |>
  filter(!is.na(.data$year))

year_topic <- gamma_df |>
  pivot_longer(starts_with("topic_"), names_to = "topic", values_to = "gamma") |>
  group_by(year, topic) |>
  summarise(mean_gamma = mean(gamma, na.rm = TRUE), .groups = "drop")
```

```{r plot-lda-faceted, fig.height=10, fig.cap="**LDA:** mean posterior topic proportion per calendar year (each panel is one topic)."}
ggplot(year_topic, aes(year, mean_gamma)) +
  geom_area(fill = "#718096", alpha = 0.35) +
  geom_line(color = "#2d3748", linewidth = 0.35) +
  facet_wrap(~topic, ncol = 4, scales = "free_y") +
  labs(
    title = "LDA topic prevalence over time (full General Conference talk corpus)",
    subtitle = paste0("K = ", params$lda_k, " | documents = all normalized talks in genconf"),
    x = NULL,
    y = "Mean gamma (topic proportion)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )
```

```{r plot-lda-heatmap, fig.height=7, fig.cap="Heatmap of z-scored yearly topic means (structure over eras)."}
zt <- year_topic |>
  group_by(topic) |>
  mutate(z = as.numeric(scale(mean_gamma))) |>
  ungroup()

ggplot(zt, aes(year, topic, fill = z)) +
  geom_tile(color = "white", linewidth = 0.08) +
  scale_fill_gradient2(
    low = "#1e3a5f", mid = "#f7fafc", high = "#9b2c2c",
    midpoint = 0, name = "z-score"
  ) +
  labs(
    title = "Topic footprint across years (z-scored mean gamma)",
    x = "Year",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_text(size = 8)
  )
```

# Reproducibility

```{r sessioninfo}
sessionInfo()
reticulate::py_config()
```
