---
title: "Prescriptive trajectory — semantic chunks & BGE tf–IDF pooling"
subtitle: "Regex-mined exemplar poles; FastEmbed-compatible weights; Parquet round-trip"
author: "conferencestats analysis"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
    number_sections: true
    df_print: paged
    fig_width: 10
    fig_height: 5.5
params:
  repo_root: !r if (nzchar(Sys.getenv("CONFERENCESTATS_ROOT", ""))) Sys.getenv("CONFERENCESTATS_ROOT") else normalizePath("..")
  # Optional: skip Python if outputs already exist (faster knit)
  skip_embed_if_present: false
  py_venv: !r file.path(normalizePath(".."), "analysis", "python", ".venv", "bin", "python")
  parquet_talks: !r file.path(normalizePath(".."), "analysis", "data", "gc_talks_normalized.parquet")
  embed_out_dir: !r file.path(normalizePath(".."), "analysis", "data", "gc_chunk_embed")
  max_chunk_tokens: 220
  batch_size: 8
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.path = "figure_presc_chunks/",
  dpi = 120
)
repo_root <- params$repo_root
if (!dir.exists(file.path(repo_root, "R"))) {
  stop("params$repo_root must point at the conferencestats package root.")
}
```

# What this report does

This is the **Parquet → Python → Parquet → R** pass you asked for:

1. **Every talk** is split into **semantic chunks** with [**semantic-text-splitter**](https://pypi.org/project/semantic-text-splitter/), using the **same HuggingFace tokenizer** as the embedding model and a **token budget** per chunk (`params$max_chunk_tokens`; default ~220 tokens, well inside BGE’s 512 context).

2. **Exemplar poles** are **real sentences from the corpus**: we scan talk text, split into sentence-like spans, and keep strings that match the **same prescriptive vs invitational regex bundles** as `general_conference_nlp_report.Rmd`. If the corpus yields too few hits, the Python script pads with short hand-authored fallbacks (see `gc_chunk_embed_pipeline.py`).

3. **Subword tf×idf weights** are computed on the **chunk collection** (each chunk = one IDF “document”), then **last-layer hidden states** are pooled and L2-normalized (**`analysis/python/hf_tfidf_pool.py`**).

4. The model is **`BAAI/bge-small-en-v1.5`**, the default **FastEmbed** ONNX family. **FastEmbed’s `embed()` API does not expose token vectors**, so **PyTorch Transformers** performs the tf–idf-weighted pooling; weights match the public HF checkpoint FastEmbed wraps.

5. For each chunk: **net = cos(chunk, U_presc) − cos(chunk, U_gentle)** with pole directions **U** = unit-mean of regex-mined (then padded) exemplar embeddings. **Per-talk score** = **mean net across chunks** (equal weight per chunk).

6. Results are **`talk_scores.parquet`** (one row per talk) and **`chunks_scored.parquet`** (diagnostics). This Rmd loads **`talk_scores.parquet`** and fits a **GAM** on year.

```{r branding, echo=FALSE, results='asis'}
cat(
  '<blockquote style="border-left:4px solid #2c5282;padding-left:1em;color:#2d3748;">',
  "Corpus: <code>generalconference::genconf</code> via <code>gc_normalize_talks()</code>. ",
  "Chunking: semantic-text-splitter + BGE tokenizer. ",
  "Embeddings: BGE-small + tf–idf-weighted pooling (HF). ",
  "</blockquote>"
)
```

---

# Prerequisites

**R**

```r
install.packages(c("pkgload", "dplyr", "ggplot2", "mgcv", "nanoparquet",
  "rmarkdown", "knitr", "remotes"))
remotes::install_github("bryanwhiting/generalconference@v0.3.2")
```

**Python (3.10+ recommended)** — create a venv under `analysis/python/` and install:

```bash
cd analysis/python
python3.11 -m venv .venv
source .venv/bin/activate   # Windows: .venv\Scripts\activate
pip install -r requirements-gc-embed.txt
```

Optional: `python gc_chunk_embed_pipeline.py --fastembed-smoke` checks that **FastEmbed** loads; encoding still uses **Transformers** for pooling.

---

# Step 1 — Export talks to Parquet

```{r export-parquet}
library(dplyr)
pkgload::load_all(repo_root, quiet = TRUE)

if (!requireNamespace("generalconference", quietly = TRUE)) {
  stop("Install generalconference v0.3.2 (see Prerequisites).")
}
if (!requireNamespace("nanoparquet", quietly = TRUE)) {
  stop("Install nanoparquet.")
}

dir.create(dirname(params$parquet_talks), recursive = TRUE, showWarnings = FALSE)
talks <- gc_normalize_talks(gc_as_tibble_generalconference())
export_tbl <- talks |> dplyr::select(talk_id, year, month, speaker, title, text)
nanoparquet::write_parquet(export_tbl, params$parquet_talks)
knitr::kable(
  utils::head(as.data.frame(export_tbl), 2),
  caption = "Exported columns (text truncated in viewer)"
)
message("Wrote ", nrow(export_tbl), " talks → ", params$parquet_talks)
```

# Step 2 — Chunk + embed (Python)

```{r run-python-pipeline}
talk_parquet <- normalizePath(params$parquet_talks, mustWork = TRUE)
out_dir <- normalizePath(params$embed_out_dir, mustWork = FALSE)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

chunk_parquet <- file.path(out_dir, "chunks_scored.parquet")
talk_score_parquet <- file.path(out_dir, "talk_scores.parquet")

run_py <- !isTRUE(params$skip_embed_if_present) ||
  !file.exists(chunk_parquet) ||
  !file.exists(talk_score_parquet)

py <- params$py_venv
if (!nzchar(py) || !file.exists(py)) {
  py <- Sys.which("python3")
}
if (!nzchar(py)) {
  stop("No Python interpreter found; set params$py_venv to venv python.")
}

pipeline_py <- normalizePath(file.path(repo_root, "analysis", "python", "gc_chunk_embed_pipeline.py"))
if (run_py) {
  args <- c(
    pipeline_py,
    "--input", talk_parquet,
    "--output-dir", out_dir,
    "--max-chunk-tokens", as.character(as.integer(params$max_chunk_tokens)),
    "--batch-size", as.character(as.integer(params$batch_size))
  )
  message("Running: ", py, " ", paste(args, collapse = " "))
  log <- system2(py, args, stdout = TRUE, stderr = TRUE)
  exit_c <- attr(log, "status")
  if (!is.null(exit_c) && exit_c != 0L) {
    stop("Python pipeline failed (exit ", exit_c, "):\n", paste(log, collapse = "\n"))
  }
  message(paste(log, collapse = "\n"))
} else {
  message("Skipping Python (outputs already present).")
}
```

# Step 3 — Load scores & model trend

```{r load-scores}
library(mgcv)
talk_sc <- nanoparquet::read_parquet(talk_score_parquet)
talk_sc <- talk_sc |>
  dplyr::filter(!is.na(.data$year), !is.na(.data$mean_net_presc)) |>
  dplyr::mutate(year = as.integer(.data$year))

knitr::kable(
  utils::head(as.data.frame(talk_sc), 5),
  caption = "Per-talk aggregate (mean net cosine across semantic chunks)"
)

m_net <- mgcv::gam(mean_net_presc ~ s(year, k = 12), data = talk_sc, method = "REML")
```

```{r plot-year, fig.cap="Talk-level mean net score vs year (semantic chunks, regex-mined exemplar poles)."}
library(ggplot2)
ggplot(talk_sc, aes(.data$year, .data$mean_net_presc)) +
  geom_hline(yintercept = 0, linetype = 3, alpha = 0.5) +
  geom_point(aes(size = .data$n_chunks), alpha = 0.15, color = "#1a365d") +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 12), color = "#c53030", fill = "#c5303033") +
  scale_size(range = c(0.3, 3), guide = "none") +
  labs(
    title = "Prescriptive vs invitational embedding drift (chunked talks)",
    subtitle = "BGE-small, tf–idf-weighted pool; poles = corpus regex exemplars",
    x = "Conference year",
    y = "Mean chunk net (cos presc − cos gentle)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

```{r plot-decade, fig.cap="Decadal means ± SE."}
by_dec <- talk_sc |>
  dplyr::mutate(decade = floor(.data$year / 10) * 10) |>
  dplyr::group_by(.data$decade) |>
  dplyr::summarise(
    n = dplyr::n(),
    m = mean(.data$mean_net_presc),
    se = stats::sd(.data$mean_net_presc) / sqrt(dplyr::n()),
    .groups = "drop"
  )

ggplot(by_dec, aes(.data$decade, .data$m)) +
  ggplot2::geom_ribbon(aes(ymin = .data$m - .data$se, ymax = .data$m + .data$se), alpha = 0.25, fill = "#2b6cb0") +
  geom_line(color = "#2c5282", linewidth = 0.9) +
  geom_point(size = 2.5, color = "#1a365d") +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  labs(title = "Decadal means (chunk-based pipeline)", x = "Decade", y = "Mean net cosine ± SE") +
  theme_minimal(base_size = 13)
```

```{r gam-note, echo=FALSE, results='asis'}
cat(
  "<p><strong>GAM</strong> <code>mean_net_presc ~ s(year)</code>: edf ≈ ",
  format(sum(m_net$edf[-1]), digits = 2),
  ", approximate <em>p</em> = ",
  format.pval(summary(m_net)$s.table[1, 4], digits = 4),
  ".</p>"
)
```

```{r interpret-tails, echo=FALSE, results='asis'}
grid_y <- seq(min(talk_sc$year), max(talk_sc$year), length.out = 400)
pdat <- data.frame(year = grid_y)
pdat$fit <- as.numeric(predict(m_net, newdata = pdat))
qlo <- stats::quantile(talk_sc$year, 0.1)
qhi <- stats::quantile(talk_sc$year, 0.9)
early <- mean(pdat$fit[pdat$year <= qlo])
late <- mean(pdat$fit[pdat$year >= qhi])
delta <- late - early
cat(
  "<p><strong>Smoothed tail comparison:</strong> later minus earlier (10% year quantiles) = ",
  "<strong>", format(delta, digits = 3), "</strong>. ",
  "Negative ⇒ average chunk semantics move toward the <em>invitational</em> pole relative to the <em>prescriptive</em> pole on this construction.</p>"
)
```

---

# Reproducibility

```{r meta, echo=FALSE, results='asis'}
meta_j <- file.path(out_dir, "pipeline_meta.json")
if (file.exists(meta_j)) {
  cat("<pre>\n", paste(readLines(meta_j, warn = FALSE), collapse = "\n"), "\n</pre>\n", sep = "")
}
```

```{r sessioninfo}
sessionInfo()
```
