---
title: "Prescriptive vs. invitational language in General Conference"
subtitle: "Semantic trajectory using exemplar poles & subword TF–IDF–weighted MiniLM pooling"
author: "conferencestats analysis"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
    number_sections: true
    df_print: paged
    fig_width: 10
    fig_height: 5.5
params:
  repo_root: !r if (nzchar(Sys.getenv("CONFERENCESTATS_ROOT", ""))) Sys.getenv("CONFERENCESTATS_ROOT") else normalizePath("..")
  minilm_max_length: 256
  tfidf_pool_batch_size: 8
  max_paragraph_rows: null
  min_paragraph_chars: 40
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.path = "figure_presc_semantic/",
  dpi = 120
)
repo_root <- params$repo_root
if (!dir.exists(file.path(repo_root, "R"))) {
  stop("params$repo_root must point at the conferencestats package root.")
}
```

# Executive summary

This report estimates whether **General Conference** rhetoric (paragraph-level) has moved toward **prescriptive / mandatory** language or toward **invitational / gentle** language over time.

**Semantic poles** are **not** mined from the corpus with regular expressions. Instead, we **hand-pick short exemplar sentences** that try to capture (a) declarative, commandment-style, warning-style teaching and (b) Christ-centered invitation, tenderness, and reflection—then average their embedding directions into two **unit reference directions** in MiniLM space.

Every **paragraph** is encoded with **`sentence-transformers/all-MiniLM-L6-v2`** using **last-layer subword pooling** whose position weights are proportional to **subword tf×idf** on the model tokenizer (same setup as `analysis/python/st_minilm_tfidf_pool.py`, IDF fit on the **paragraph corpus** so scoring stays anchored to conference language).

For each paragraph we compute

> **net semantic prescriptiveness** = cos(paragraph, pole_prescriptive) − cos(paragraph, pole_invitational)

and aggregate to **per-talk** means, then smooth **against conference year** with a **GAM**.

**How to read the rest:** Methods spell out exemplars and pooling; Results give plots and tables; **Interpretation** states the directional answer (including whether recent years might tilt slightly back toward prescriptive language if the curve is non-monotone).

```{r branding, echo=FALSE, results='asis'}
cat(
  '<blockquote style="border-left:4px solid #2c5282;padding-left:1em;color:#2d3748;">',
  "<strong>Corpus:</strong> <code>generalconference::genconf</code> (paragraph rows joined to normalized talks). ",
  "<br/><strong>Model & pooling:</strong> ",
  '<a href="https://huggingface.co/sentence-transformers/all-MiniLM-L6-v2">all-MiniLM-L6-v2</a>',
  " + tf–idf-weighted subword pool.",
  "</blockquote>"
)
```

---

# Prerequisites

**R** (once):

```r
install.packages(c(
  "pkgload", "rmarkdown", "knitr", "reticulate", "dplyr", "tidyr",
  "stringr", "ggplot2", "scales", "mgcv", "tibble", "remotes"
))
remotes::install_github("bryanwhiting/generalconference@v0.3.2")
```

**Python** (CPU or GPU; first run downloads weights):

```{r py-bootstrap, eval=FALSE}
reticulate::py_install(c("transformers", "torch", "numpy"), pip = TRUE)
```

```{r py-ensure}
library(reticulate)
have_tf <- tryCatch({ import("transformers"); TRUE }, error = function(e) FALSE)
if (!have_tf) {
  message("Installing transformers + torch via pip (one-time, can take several minutes)...")
  py_install(c("transformers", "torch"), pip = TRUE)
}
```

---

# Data: talks and paragraphs

```{r load-pkg}
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(mgcv)
library(tibble)
pkgload::load_all(repo_root, quiet = TRUE)

if (!requireNamespace("generalconference", quietly = TRUE)) {
  stop("Install generalconference v0.3.2 (see Prerequisites).")
}

talks <- gc_normalize_talks(gc_as_tibble_generalconference())
stopifnot(nrow(talks) > 100L)

talks |> 
  summarise(
    n_talks = dplyr::n(),
    year_min = min(year, na.rm = TRUE),
    year_max = max(year, na.rm = TRUE),
    n_words = sum(word_count, na.rm = TRUE)
  ) |>
  knitr::kable(caption = "Corpus coverage")
```

```{r paragraphs}
para <- gc_as_paragraph_tibble_generalconference() |>
  inner_join(
    talks |>
      select(talk_id, year, month, conference_date, speaker, title, url, word_count),
    by = c("year", "month", "speaker", "title", "url")
  ) |>
  filter(nchar(.data$text) >= params$min_paragraph_chars)

if (!is.null(params$max_paragraph_rows) && is.numeric(params$max_paragraph_rows) &&
  params$max_paragraph_rows > 0L) {
  set.seed(42)
  para <- dplyr::slice_sample(para, n = min(nrow(para), as.integer(params$max_paragraph_rows)))
}

knitr::kable(
  utils::head(as.data.frame(para), 3),
  caption = "Sample paragraph rows (text truncated in viewer)"
)
message("Paragraphs encoded: ", nrow(para))
```

---

# Methods

## Exemplar poles (hand-authored)

The lists below are **deliberate prototypes**, not corpus-derived. You should edit them to match your research intent. Two sentences mirror prompts you gave for “prescriptive / doctrinal statement” flavor; spelling (“Malachy”) is kept in one exemplar as in the prompt, and a parallel line uses “Malachi” for a cleaner tokenization.

```{r exemplars, echo=TRUE}
exemplar_prescriptive <- c(
  "I should put priesthood keys of the seeming power upon Peter, James and John.",
  "It is also significant that when Christ visited the people on the American continent he specifically recited to them the Old Testament account in Malachy.",
  "It is also significant that when Christ visited the people on the American continent he specifically recited to them the Old Testament account in Malachi.",
  "Thou shalt have no other gods before me.",
  "Thou shalt not commit adultery.",
  "Repent ye, for the kingdom of heaven is at hand.",
  "Ye must press forward with a steadfastness in Christ.",
  "Do not procrastinate the day of your repentance.",
  "Beware of pride, for it will destroy your soul.",
  "Woe unto him who rebels against the Lord.",
  "Cease from all ungodliness and forsake your sins."
)

exemplar_invitational <- c(
  "I invite you to come unto Christ and be healed.",
  "We extend a heartfelt invitation to each of you to feel His redeeming love.",
  "Ponder the tender mercies the Lord has shown you.",
  "Consider how gentle the Savior is with His children.",
  "His mercy and peace rest upon all who turn to Him.",
  "We invite you to receive the peace that only He can give.",
  "Jesus Christ is our Savior; yield your heart to Him in love.",
  "As we invite the Spirit, our hearts are softened.",
  "Treasure up these words and let them nourish your faith.",
  "Come unto Christ, and lay your burdens at His feet."
)

tibble::tibble(side = "prescriptive", text = exemplar_prescriptive) |>
  bind_rows(tibble::tibble(side = "invitational", text = exemplar_invitational)) |>
  knitr::kable(caption = "Exemplar strings defining the two poles")
```

## Subword tf–idf pooling (MiniLM)

1. **IDF** is computed over **paragraph strings only** (each paragraph = one “document” for df). Token ids appearing in CLS/SEP/PAD/UNK are excluded from df updates, matching `compute_subword_idf`.
2. Each span (paragraph or exemplar) is run through **`AutoModel` last hidden states**.
3. For each non-special subword position \(j\) with type \(t_j\), weight \(w_j \propto \text{tf}(t_j)\cdot \text{idf}(t_j)\) within that span, normalized to sum to 1.
4. Pooled vector = \(\sum_j w_j h_j\), then **L2-normalized**.
5. **Pole vectors**: encode all exemplar strings with the **same** paragraph IDF; take the **mean** of exemplar rows and L2-normalize (`unit_mean_direction` in Python).

```{r connect-python}
py_dir <- normalizePath(file.path(repo_root, "analysis", "python"))
sys <- reticulate::import("sys", convert = FALSE)
sys$path$insert(0L, py_dir)
stw <- reticulate::import("st_minilm_tfidf_pool", convert = TRUE)

ml <- as.integer(params$minilm_max_length)
tbs <- params$tfidf_pool_batch_size
bs_pool <- if (is.numeric(tbs) && length(tbs) == 1L && !is.na(tbs) && tbs > 0) {
  as.integer(tbs)
} else {
  8L
}

txt <- as.character(para$text)
message("Fitting subword IDF on ", length(txt), " paragraphs...")
idf <- stw$compute_subword_idf(txt, max_length = ml)

message("Encoding paragraphs (tf–idf pool)...")
emb <- stw$encode_tfidf_weighted_pool(txt, idf, max_length = ml, batch_size = bs_pool)
stopifnot(is.matrix(emb), nrow(emb) == nrow(para), ncol(emb) == 384L)

all_ex <- c(exemplar_prescriptive, exemplar_invitational)
message("Encoding exemplars...")
emb_ex <- stw$encode_tfidf_weighted_pool(all_ex, idf, max_length = ml, batch_size = bs_pool)
np <- length(exemplar_prescriptive)
ng <- length(exemplar_invitational)

u_presc <- stw$unit_mean_direction(emb_ex[seq_len(np), , drop = FALSE])
u_gent <- stw$unit_mean_direction(emb_ex[np + seq_len(ng), , drop = FALSE])
if (is.null(u_presc) || is.null(u_gent)) {
  stop("Exemplar centroid failed — check exemplar_prescriptive / exemplar_invitational.")
}

cos_p <- as.numeric(stw$cosine_to_rows(emb, u_presc))
cos_g <- as.numeric(stw$cosine_to_rows(emb, u_gent))

para_scored <- para |>
  dplyr::mutate(
    cos_presc = cos_p,
    cos_gentle = cos_g,
    net_presc = .data$cos_presc - .data$cos_gentle
  )
```

---

# Results

## Talk-level scores and smooth trend

```{r aggregates}
talk_sc <- para_scored |>
  group_by(talk_id, year, month, conference_date, speaker) |>
  summarise(
    n_para = dplyr::n(),
    mean_net = mean(.data$net_presc, na.rm = TRUE),
    .groups = "drop"
  )

m_net <- mgcv::gam(mean_net ~ s(year, k = 12), data = talk_sc, method = "REML")
```

```{r plot-year, fig.cap="Each talk: mean paragraph net score (prescriptive pole minus invitational pole). Points sized by paragraph count; red curve = GAM smooth."}
ggplot(talk_sc, aes(year, mean_net)) +
  geom_hline(yintercept = 0, linetype = 3, alpha = 0.5) +
  geom_point(aes(size = n_para), alpha = 0.12, color = "#1a365d") +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 12), color = "#c53030", fill = "#c5303033") +
  scale_size(range = c(0.4, 3), guide = "none") +
  labs(
    title = "Semantic prescriptiveness over conference years",
    subtitle = "Exemplar-defined poles | tf–idf weighted MiniLM pooling | mean net cosine per talk",
    x = "Conference year",
    y = "Mean (cos prescriptive − cos invitational)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

```{r plot-decade, fig.cap="Decadal means of talk-level net score (error bars = standard error of the mean)."}
by_dec <- talk_sc |>
  mutate(decade = floor(year / 10) * 10) |>
  group_by(decade) |>
  summarise(
    n = dplyr::n(),
    m = mean(mean_net),
    se = sd(mean_net) / sqrt(dplyr::n()),
    .groups = "drop"
  )

ggplot(by_dec, aes(decade, m)) +
  geom_ribbon(aes(ymin = m - se, ymax = m + se), alpha = 0.25, fill = "#2b6cb0") +
  geom_line(color = "#2c5282", linewidth = 0.9) +
  geom_point(size = 2.5, color = "#1a365d") +
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5) +
  labs(
    title = "Decadal averages (exemplar poles)",
    x = "Decade",
    y = "Mean net cosine ± SE"
  ) +
  theme_minimal(base_size = 13)
```

```{r gam-note, echo=FALSE, results='asis'}
cat(
  "<p><strong>GAM</strong> <code>mean_net ~ s(year)</code>: ",
  "edf ≈ ",
  format(sum(m_net$edf[-1]), digits = 2),
  ", approximate <em>p</em> = ",
  format.pval(summary(m_net)$s.table[1, 4], digits = 4),
  ".</p>"
)
```

## Is there a “dip then rise”? (simple curvature check)

```{r curvature}
## Global quadratic in calendar year (not a GAM — quick curvature screen)
yr0 <- min(talk_sc$year)
talk_sc <- talk_sc |> mutate(year_c = year - yr0)
m_quad <- lm(mean_net ~ poly(year_c, 2, raw = FALSE), data = talk_sc)

dec_compare <- talk_sc |>
  mutate(era = cut(
    year,
    breaks = c(-Inf, 1989, 2009, Inf),
    labels = c("1971–1989", "1990–2009", "2010–2021"),
    right = TRUE
  )) |>
  group_by(era) |>
  summarise(
    n_talks = dplyr::n(),
    mean_net = mean(mean_net),
    se = sd(mean_net) / sqrt(dplyr::n()),
    .groups = "drop"
  )

coef_sq <- as.data.frame(stats::coef(summary(m_quad)))
colnames(coef_sq) <- c("estimate", "se", "t", "p")
coef_sq$term <- rownames(coef_sq)
knitr::kable(
  coef_sq |> dplyr::filter(term != "(Intercept)"),
  digits = 4,
  caption = "Polynomial screen: orthogonal quadratic in year (reference = corpus start year)"
)

knitr::kable(dec_compare, digits = 3, caption = "Mean net score by broad era (edit breaks as needed)")
```

```{r pred-compare, echo=FALSE, results='asis'}
grid_y <- seq(min(talk_sc$year), max(talk_sc$year), length.out = 400)
pdat <- tibble(year = grid_y)
pdat$fit <- as.numeric(predict(m_net, newdata = pdat))

qlo <- stats::quantile(talk_sc$year, 0.1)
qhi <- stats::quantile(talk_sc$year, 0.9)
early <- mean(pdat$fit[pdat$year <= qlo])
late <- mean(pdat$fit[pdat$year >= qhi])
delta <- late - early

cat(
  "<p><strong>GAM smooth — tails of the corpus window:</strong><br/>",
  "Approx. mean fitted net score in the <em>earlier</em> extreme (~lowest 10% of years): <strong>",
  format(early, digits = 3),
  "</strong><br/>",
  "Approx. mean fitted net score in the <em>later</em> extreme (~highest 10% of years): <strong>",
  format(late, digits = 3),
  "</strong><br/>",
  "Difference (later − earlier): <strong>",
  format(delta, digits = 3),
  "</strong> (negative ⇒ corpus drifts toward the invitational pole on this scale).</p>"
)
```

---

# Interpretation

This section restates findings in plain language using the **objects computed above** (knit the document to fill numbers).

- **Directional drift:** Compare the GAM-smoothed **early vs late** sentence in the **pred-compare** chunk. If **later − earlier** is **negative**, average paragraph semantics across talks shift toward **higher similarity to the invitational exemplar set** relative to the **prescriptive exemplar set**—as operationalized here, that reads as **less “prescriptive pole” emphasis over time**, not more.

- **“Less then more?”** If the **decadal plot** or **era table** shows the **most recent** decade **tick upward** compared with the 2000s while the **long-run GAM** still slopes down, you can describe a **modest recent partial rebound** on top of a **broad downward** trend (or the opposite pattern if the numbers go the other way). The **quadratic lm** table is only a coarse global curvature screen; the **GAM curve** is the primary visual for local reversals.

- **Caveats:** (1) Results **depend on the exemplar sentences**; swap them and the poles move. (2) **MiniLM** compresses nuance; this is **relative geometry**, not theology. (3) **Paragraph boundaries** come from the `generalconference` source; different segmentation changes scores.

---

# Next steps (outside this report)

For **semantic chunking** of whole talks, **BGE-small** (FastEmbed-compatible weights), **regex-mined exemplar poles**, and **Parquet → R**, see **`analysis/prescriptive_chunks_embed_report.Rmd`** (`analysis/python/gc_chunk_embed_pipeline.py`).

---

# Reproducibility

```{r sessioninfo}
sessionInfo()
reticulate::py_config()
```
